An Efficient Feedback Calibration Algorithm for Direct Imaging Radio Telescopes

### 2016-06-22 cal_demo/ms.tex
Bolded changes in cal demo paper in prep for resubmission

% Abstract of the paper
\begin{abstract}
We present the E-field Parallel Imaging Calibration (EPICal) algorithm, which addresses the 
need for a real-time calibration method for direct imaging radio astronomy correlators. Direct 
imaging involves a spatial fast Fourier transform of antenna signals, alleviating the harsh 
$\mathcal{O}(\Nant^2)$ computational scaling to a more gentle $\mathcal{O}(\Nant \log_2 
\Nant)$, which can save orders of magnitude in computation cost for next generation arrays 
consisting of hundreds to thousands of antennas. However, because signals are mixed in the 
correlator, gain correction must be applied on the front end. We develop the EPICal algorithm 
to form gain solutions in real time without ever forming visibilities. This method scales as the 
number of antennas, and produces results comparable to those from visibilities. Through 
simulations and application to Long Wavelength Array data we show this algorithm is a 
promising solution for next generation instruments.
\end{abstract}

@@

The most common radio correlator designs of today, the FX and lag correlators \citep{rom99},
cross-multiply the signals from all pairs of antennas, and 
the computation scales as the number of antennas squared, $\mathcal{O}(\Nant^2)$ 
\citep{bun04}. As the number of elements in future arrays grows, the computational cost will 
become prohibitively expensive, and exploring efficient correlator schemes is essential to 
enable next generation instruments \citep{lon00}. Meanwhile, radio transient monitoring 
requires access to high time and frequency resolution data to identify and characterize events 
such as fast radio bursts (FRBs, \citealt{lor07}), or to follow up gravitational wave candidates 
with radio observations \citep{abb16a,abb16b}. FRBs are highly unexplored at low frequencies 
(< 1 GHz), but are expected to occur on timescales $\Delta t \sim$ 1--10~ms \citep{tho13}. 
Recording the full visibility matrix for $\Nant \gtrsim 10^3$ arrays at this timescale leads to 
extremely high data write rates.

Direct imaging correlators are a new variety of radio correlator which aim to alleviate both the 
computational strain of forming $\Nant^2$ correlations and the high data throughput associated 
with short timescale science. This is done by performing a spatial fast Fourier transform (FFT) 
to image the antenna signals, then squaring and averaging in time. This process scales as 
$\mathcal{O}(\Ng \log_2 \Ng)$, where \Ng~is the number of grid points in the FFT \citep{mor11,
 teg09, teg10}. For certain classes of telescopes, significantly those envisioned for next 
 generation cosmology experiments, this scaling is a large improvement over the $\Nant^2$ 
 scaling of traditional methods. Furthermore, because images are generated in real-time, the native 
 output bandwidth will be lowered in cases where $\Ng < \Nant^2$, and has the potential to be 
 lowered even further with real-time transient processing.

@@

A challenge common to all direct imaging algorithms is calibration of the antenna gains. With a 
traditional cross correlator, pair-wise visibilities are written to disk and used to calibrate 
after observing before further processing such as imaging. However, a direct imaging correlator mixes 
the signals from all antennas before averaging and writing to disk, making calibration a 
requirement at the front end, before imaging and averaging. Previous solutions have involved 
applying calibration solutions generated from a parallel FX correlator \citep{zhe14, fos14}, or 
integrating a dedicated FX correlator which periodically formed the full visibility matrix to solve 
for gains \citep{wij09,dev09}. While these solutions were sufficient to enable the exploration of 
FFT correlators and beamformers, they will not scale to future arrays with $\Nant \gtrsim 10^3$.

@@

We review the MOFF algorithm and derive the calibration algorithm in \S \ref{sec:math}. We 
then demonstrate the algorithm in simulations in \S \ref{sec:sim}, 
and discuss noise trends in \S \ref{sec:noise}. We then apply the algorithm to a sample LWA 
data set in \S \ref{sec:data}.
Finally we conclude and discuss potential extensions to the 
algorithm in \S \ref{sec:discussion}.

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{fig1.pdf}
\caption{Data flow for the MOFF algorithm, reproduced from \citealt{thy15c}. The time ordered 
electric field measurements are collected by the antenna elements then frequency transformed in the 
F-engine. The calibration is applied, then the electric field measurements are gridded using the 
antenna aperture illumination patterns. The imager performs a spatial 2D FFT on the gridded data before 
squaring and averaging. The accumulated images are then written to disk.}
\label{fig:moff_flow}
\end{center}
\end{figure}

%%% New text
We begin by considering the electric field vector incident on the ground as a function of time, 
$\mathbf{E}(\mathbf{r},t)$. The position vector, $\mathbf{r}$, is measured from some arbitrary 
origin. Following \citealt{cla99} we make a few assumptions in order to clarify notation and 
simplify the problem. First we note that the time ordered electric field can be expressed as a 
Fourier series with coefficients corresponding to frequencies $f$. This allows us to analyze the 
quasi-monochromatic components of the electric field, $\mathbf{E}(\mathbf{r},f,t)$, with the 
understanding that the full field is composed of all components and their corresponding Fourier 
functions. We retain the time dependence because in practice each coefficient will only be 
measured over some sampling period.

It is convenient to simplify notation by expressing the position vector in units of wavelengths, 
$\mathbf{r}\rightarrow \mathbf{r}/\lambda$. We will also assume all antennas lie in a plane, 
reducing the problem to two dimensions. While the MOFF algorithm in principle can account for 
non-coplanar arrays through the W-projection algorithm \citep{cor08}, the EPIC code does not 
currently support this so we leave it to future work.

Next we treat the electric field as a scalar quantity, ignoring polarization effects. This allows us 
to express the electric field as a scalar, $E(\mathbf{r},f,t)$, and treat subsequent field 
propagation and measurements as scalar rather than tensor multiplications. A full polarization 
treatment is required to construct polarized sky images, however the calibration will only 
require a sky model expressed in the instrumental polarization basis. We can therefore treat 
each polarization independently.

Finally we assume the sources of the electric field are far away, allowing us to describe the 
electric fields propagating from sky sources as originating from an imaginary celestial sphere 
located a large distance from the ground. We express this intermediary electric field as an 
angular distribution, $\mathcal{E}(\hat{\mathbf{s}},f,t)$, where $\hat{\mathbf{s}}$ is the unit 
vector in the direction of a patch of the celestial sphere. We further assume the space between 
the celestial sphere and the ground is empty, allowing us to express the electric field incident 
on the ground as a Fourier integral of the propagated fields across the hemisphere above the horizon.
\begin{equation}\label{eq:sky_propagated}
E(\mathbf{r},f,t) = \int \mathcal{E}(\hat{\mathbf{s}},f,t) e^{-2\pi i \mathbf{r} \cdot \hat{\mathbf{s}}}\, d\Omega
\end{equation}  
By making the latter assumptions we have left treatment of the ionosphere and terrestrial 
emission such as local interference or thermal radiation from the ground for future work.

Next we turn to the measurement performed by an antenna with far-field radiation pattern 
$\mathcal{W}_a(\hat{\mathbf{s}})$, where the subscript $a$ denotes a specific antenna. We 
adopt the common convention of normalizing the radiation pattern to have response of 1 at the 
center of the main lobe \citep[e.g.][]{nap99}. Consequently, the translation between the terminal 
voltages to electric field measurements has been absorbed in our definition of the radiation 
pattern. The ideal antenna electric field measurement can be expressed as 
\begin{equation}
E^I_a(f,t) = \int \mathcal{W}_a(\hat{\mathbf{s}})\mathcal{E}(\hat{\mathbf{s}},f,t) e^{-2\pi i \mathbf{r} \cdot \hat{\mathbf{s}}}\, d\Omega. 
\end{equation}
Equivalently we can express the effect of the antenna radiation pattern in the plane of the 
aperture.
\begin{equation}
E^I_a(f,t) = \int_{\mathrm{aperture}}W_a(\mathbf{r}-\mathbf{r}_a)E_a(\mathbf{r},f,t)\,d^2\mathbf{r}
\end{equation}
Here we used the Fourier relationship between the antenna's far-field radiation pattern and the 
aperture illumination pattern (sometimes referred to as the aperture distribution), $W_a(\mathbf{r}-\mathbf{r}_a)=\int  \mathcal{W}_a(\hat{\mathbf{s}}) \, d
\Omega$. 
While it is possible to define each antenna aperture illumination pattern in a way to absorb the 
relative position of the antenna, we keep the $\mathbf{r}_a$ coordinate to retain the separable
concepts of the primary beam (the product of two antenna radiation patterns) and the array
point spread function (square of the Fourier transform of the array layout).

%The electric field incident on the ground, $\Er{}(\mathbf{r},t)$, is related to the sky electric field, 
%$E(\hat{\mathbf{s}},t)$, through a Fourier transform.
%\begin{equation}
%\Er{}(\mathbf{r},t) = \int E(\hat{\mathbf{s}},t) e^{-2\pi i \mathbf{r}\cdot \hat{\mathbf{s}}}\, \dif^2 \hat{\mathbf{s}}
%\end{equation}
%Here $\hat{\mathbf{s}}$ denotes the sine-projected unit vector for the sky angle, $\mathbf{r}$ is 
%the observer's location (measured in wavelengths relative to an arbitrary origin). We will 
%encounter several quantities which we attempt to estimate. We distinguish the ``true'' values 
%with a superscript $T$, while the estimated quantities are denoted with a prime. We define the 
%true antenna signal as a convolution of the antenna voltage pattern, \beamr, with the electric 
%field on the ground.
%\begin{equation}
%\Er{a}^T(t) \equiv \int \beamr_a(\mathbf{r}-\ra) \Er{}(\mathbf{r},t) \, \dif^2 \mathbf{r}
%\end{equation}
%The subscript $a$ labels the antenna, and $\mathbf{r}_a$ is the location of antenna $a$.

The first step of the MOFF algorithm is to digitize the antenna signals and perform an FFT to 
channelize the time-ordered data. This is effectively estimating the quasi-monochromatic 
components of the electric field we already discussed. The sampling rate of the digitization 
determines the bandwidth of the observation, and the sampling period determines the minimum 
frequency channel resolution. The frequency Fourier transform operation is often referred to as 
the F-engine of a correlator. The MOFF shares the same F-engine design as an FX cross
correlator which performs the frequency transform (F-engine) before cross multiplying 
(X-engine).

%The first step of the MOFF algorithm is identical to the F-engine of a traditional FX correlator 
%where the electric fields are Fourier transformed to frequency channels, $\Er{a}^T(f,t)$. Even 
%after transforming along the time axis, we retain the time dependence of the signal to represent 
%a stream of spectra.

Thus far we have considered the perfectly measured electric field distribution according to 
the antenna far-field radiation patterns. Here we introduce a multiplicative complex gain as 
well as an additive noise term which corrupt the measured signals.
\begin{equation}\label{eq:apply_gain}
E_a(f,t) = g_a(f,t) E_a^I(f,t) + n_a(f,t)
\end{equation}
%The noise term is 
%strictly receiver noise -- noise introduced by the instrument. Any sky noise is implicitly included 
%in the time dependence of the sky electric field. 
We have carried the frequency dependence to this point to make clear the complex gains are
frequency dependent. However, each subsequent step treats each frequency 
channel independently, so we will drop the $f$ to simplify notation. 
Furthermore, while antenna gains may vary with time, we will assume they are constant over
the period of our calibration, and thus drop their time dependence from the notation.

The MOFF algorithm next calls for a calibration of the electric field measurements. The goal of our new calibration method will be 
to form an estimate of the gains, $g'_a$, where the prime represents an estimate. For now we will assume we have formed an estimate 
to proceed with the MOFF pipeline. We correct the incoming electric field data stream using our 
current estimate of the gains.
\begin{equation}
E'_a(t) = E_a(t)/ g'_a
\end{equation}

The next step is to grid the calibrated electric field measurements to a finite set of regularly spaced 
positions on the 2D plane of the ground, $\mathbf{r}_j$.
The extent of the grid is chosen to encompass the footprint of the antenna array.
MOFF uses the antenna aperture illumination patterns as the gridding 
kernel according to the software holography/A-transpose technique \citep{mor09,bha08}. 
Mathematically we can express the gridding operation as
\begin{equation}\label{eq:gridding}
E'_j=\sum_a W_a(\mathbf{r}_j-\mathbf{r}_a) E'_a(t)
\end{equation}
Formally each pixel is calculated by summing over all antennas and their aperture illumination patterns.
However the aperture illumination patterns are typically compact and non-overlapping, so most pixels
will only contain contribution from at most one antenna, allowing equation~\ref{eq:gridding}
to be calculated through a sparse matrix multiplication.
This gridding 
achieves an optimal map \citep{teg97b} and places the data on a regular grid. Unlike 
traditional correlators which grid visibilities, the MOFF grids the electric field measurements directly. This 
operation must be performed for every time sample, before any averaging of the data.

After gridding, the imaging portion of the MOFF performs a 2D spatial FFT to form 
instantaneous electric field images of the sky at fixed locations $\s_k$.
The Fourier dual coordinates to $\mathbf{r}$ are $(l,m)$, where $l=\sin\theta\cos\phi$,
$m=\nobreak\sin\theta\sin\phi$, and $(\theta,\phi)$ are the zenith 
and azimuthal angles. The full unit vector towards the sky is 
$\s=(l,m,\sqrt{1-l^2-m^2})$.
The sample spacing and resolution of $\s_k$ are given 
by the choices of $\mathbf{r}_j$ grid parameters, according to the usual FFT relations. 
Fine $\mathbf{r}_j$ grid spacing can result in non-physical angular pixels ($l^2+m^2>1$),
which should be discarded.

The pipeline up to this point can be summarized 
through the following equation.
\begin{align}
\mathcal{E}'(\s_k,t) & = \underbrace{\frac{1}{\Nant} \sum_j e^{2\pi i \mathbf{r}_j \cdot \s_k}}_{\mathrm{2D\;FFT}} 
\underbrace{\sum_a W_a(\mathbf{r}_j - \ra)}_{\mathrm{Gridding}} \\
& \qquad \times \underbrace{ \frac{1}{g'_a} \left(g_a E_a^I(t)+n_a(t)\right)}_{\mathrm{Calibration}} \nonumber
\end{align}
The sum over antennas uses the antenna aperture illumination pattern to grid the calibrated electric field measurements onto 
regular gridpoints, $\mathbf{r}_j$. The sum over $j$ denotes the 2D FFT to sky coordinates, resulting in an 
estimate for the instantaneous electric field image. We can simplify this expression by 
exchanging the sums and use the DFT to approximate the Fourier relationship between
the antenna aperture illumination pattern and the far-field radiation pattern.
\begin{align}\label{eq:epix}
\mathcal{E}'(\s_k,t) & = \frac{1}{\Nant} \sum_a \frac{1}{g'_a}\left(g_a E^I_a(t)+n_a(t)\right) e^{2\pi i \s_k \cdot \ra} \nonumber \\
  & \qquad\times \sum_j W_a(\mathbf{r}_j-\ra)e^{2\pi i \s_k \cdot (\mathbf{r}_j-\ra)} \nonumber\\
& \approx \frac{1}{\Nant} \sum_a \frac{1}{g'_a}\left(g_aE^I_a(t)+n_a(t)\right) e^{2\pi i \s_k \cdot \ra}\mathcal{W}_a(\s_k)
\end{align}
In this form we see that the effect of gridding with the antenna aperture illumination pattern is to attenuate 
the image by a factor of the far-field radiation response, $\mathcal{W}_a(\s)$.

Finally, the images are squared and averaged in time to form dirty images, 
$I'(\hat{\mathbf{s}}_k)$, then written 
to disk. While operations have been reordered, the resulting image is equivalent to one 
produced by gridding and imaging visibilities. By shifting from cross-correlation of all pairs of 
antennas to a spatial FFT, the computational cost is significantly lowered for certain classes of 
compact arrays. The output bandwidth is also greatly reduced as visibilities are exchanged for 
gridded images. A full exploration of these scalings was presented in \citealt{thy15c}.

@@

As a starting point, we consider the feedback calibration outlined in \citealt{mor11}. There it 
was suggested to form a correlation of the uncalibrated antenna measurements with the complex conjugate of an image 
pixel from the output of the correlator. This is a statistically stationary quantity, which can be 
related to the sum of visibilities involving the antenna used in the correlation -- exactly the sum 
needed to calibrate a simple sky of a single point source. However, we aim for a more 
generalized solution for arbitrarily complex sky models. We therefore study the full expression 
for the antenna-pixel correlation,
\begin{equation}\label{eq:Kna_def}
\Kna \equiv \left<E_a^{(n)}(t) \mathcal{E}'^{*(n)}(\spix,t)\right>_t,
\end{equation}
where the superscript $n$ again represents the quantity formed in the $n^\mathrm{th}$ 
calibration loop, 
the superscript $^*$ indicates complex conjugation,
and \spix\, is the pixel center nearest a bright calibrator of interest. The 
following will hold for any chosen pixel, \spix, though it is advantageous to choose a pixel which 
contains a bright source to achieve a high signal to noise.

>>>>

% Abstract of the paper
\begin{abstract}
We present the E-field Parallel Imaging Calibration (EPICal) algorithm, which addresses the 
need for a real-time calibration method for direct imaging radio astronomy correlators. Direct 
imaging involves a spatial fast Fourier transform of antenna \textbf{signals}, alleviating the harsh 
$\mathcal{O}(\Nant^2)$ computational scaling to a more gentle $\mathcal{O}(\Nant \log_2 
\Nant)$, which can save orders of magnitude in computation cost for next generation arrays 
consisting of hundreds to thousands of antennas. However, because signals are mixed in the 
correlator, gain correction must be applied on the front end. We develop the EPICal algorithm 
to form gain solutions in real time without ever forming visibilities. This method scales as the 
number of antennas, and produces results comparable to those from visibilities. Through 
simulations and application to Long Wavelength Array data we show this algorithm is a 
promising solution for next generation instruments.
\end{abstract}

@@

\textbf{The most common radio correlator designs of today, the FX and lag correlators \citep{rom99}},
cross-multiply the \textbf{signals} from all pairs of antennas, and 
the computation scales as the number of antennas squared, $\mathcal{O}(\Nant^2)$ 
\citep{bun04}. As the number of elements in future arrays grows, the computational cost will 
become prohibitively expensive, and exploring efficient correlator schemes is essential to 
enable next generation instruments \citep{lon00}. Meanwhile, radio transient monitoring 
requires access to high time and frequency resolution data to identify and characterize events 
such as fast radio bursts (FRBs, \citealt{lor07}), or to follow up gravitational wave candidates 
with radio observations \citep{abb16a,abb16b}. FRBs are highly unexplored at low frequencies 
(< 1 GHz), but are expected to occur on timescales $\Delta t \sim$ 1--10~ms \citep{tho13}. 
Recording the full visibility matrix for $\Nant \gtrsim 10^3$ arrays at this timescale leads to 
extremely high data write rates.

Direct imaging correlators are a new variety of radio correlator which aim to alleviate both the 
computational strain of forming $\Nant^2$ correlations and the high data throughput associated 
with short timescale science. This is done by performing a spatial fast Fourier transform (FFT) 
to image the antenna signals, then squaring and averaging in time. This process scales as 
$\mathcal{O}(\Ng \log_2 \Ng)$, where \Ng~is the number of grid points in the FFT \citep{mor11,
 teg09, teg10}. For certain classes of telescopes, significantly those envisioned for next 
 generation cosmology experiments, this scaling is a large improvement over the $\Nant^2$ 
 scaling of traditional methods. Furthermore, because images are generated \textbf{in real-time}, the native 
 output bandwidth will be lowered \textbf{in cases where $\Ng < \Nant^2$}, and has the potential to be 
 lowered even further with \textbf{real-time} transient processing.

@@

A challenge common to all direct imaging algorithms is calibration of the antenna gains. With a 
traditional \textbf{cross correlator}, pair-wise visibilities are written to disk and used to calibrate 
\textbf{after observing before further processing} such as imaging. However, a direct imaging correlator mixes 
the signals from all antennas before averaging and writing to disk, making calibration a 
requirement at the front end, before imaging and averaging. Previous solutions have involved 
applying calibration solutions generated from a parallel FX correlator \citep{zhe14, fos14}, or 
integrating a dedicated FX correlator which periodically formed the full visibility matrix to solve 
for gains \citep{wij09,dev09}. While these solutions were sufficient to enable the exploration of 
FFT correlators and beamformers, they will not scale to future arrays with $\Nant \gtrsim 10^3$.

@@

We review the MOFF algorithm and derive the calibration algorithm in \S \ref{sec:math}. We 
then demonstrate the algorithm in simulations in \S \ref{sec:sim}, 
\textbf{and discuss noise trends in \S \ref{sec:noise}. We then apply the algorithm to a sample LWA 
data set in \S \ref{sec:data}}.
Finally we conclude and discuss potential extensions to the 
algorithm in \S \ref{sec:discussion}.

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{fig1.pdf}
\caption{Data flow for the MOFF algorithm, reproduced from \citealt{thy15c}. The time ordered 
electric field \textbf{measurements} are collected by the antenna elements then frequency transformed in the 
F-engine. The calibration is applied, then the electric field measurements are gridded using the 
\textbf{antenna aperture illumination patterns}. The imager performs a spatial 2D FFT on the gridded data before 
squaring and averaging. The accumulated images are then written to disk.}
\label{fig:moff_flow}
\end{center}
\end{figure}

%%% New text
\textbf{We begin by considering the electric field vector incident on the ground as a function of time, 
$\mathbf{E}(\mathbf{r},t)$. The position vector, $\mathbf{r}$, is measured from some arbitrary 
origin. Following \citealt{cla99} we make a few assumptions in order to clarify notation and 
simplify the problem. First we note that the time ordered electric field can be expressed as a 
Fourier series with coefficients corresponding to frequencies $f$. This allows us to analyze the 
quasi-monochromatic components of the electric field, $\mathbf{E}(\mathbf{r},f,t)$, with the 
understanding that the full field is composed of all components and their corresponding Fourier 
functions. We retain the time dependence because in practice each coefficient will only be 
measured over some sampling period. }

\textbf{It is convenient to simplify notation by expressing the position vector in units of wavelengths, 
$\mathbf{r}\rightarrow \mathbf{r}/\lambda$. We will also assume all antennas lie in a plane, 
reducing the problem to two dimensions. While the MOFF algorithm in principle can account for 
non-coplanar arrays through the W-projection algorithm \citep{cor08}, the EPIC code does not 
currently support this so we leave it to future work.}

\textbf{Next we treat the electric field as a scalar quantity, ignoring polarization effects. This allows us 
to express the electric field as a scalar, $E(\mathbf{r},f,t)$, and treat subsequent field 
propagation and measurements as scalar rather than tensor multiplications. A full polarization 
treatment is required to construct polarized sky images, however the calibration will only 
require a sky model expressed in the instrumental polarization basis. We can therefore treat 
each polarization independently.}

\textbf{Finally we assume the sources of the electric field are far away, allowing us to describe the 
electric fields propagating from sky sources as originating from an imaginary celestial sphere 
located a large distance from the ground. We express this intermediary electric field as an 
angular distribution, $\mathcal{E}(\hat{\mathbf{s}},f,t)$, where $\hat{\mathbf{s}}$ is the unit 
vector in the direction of a patch of the celestial sphere. We further assume the space between 
the celestial sphere and the ground is empty, allowing us to express the electric field incident 
on the ground as a Fourier integral of the propagated fields across the hemisphere above the horizon.
\begin{equation}\label{eq:sky_propagated}
E(\mathbf{r},f,t) = \int \mathcal{E}(\hat{\mathbf{s}},f,t) e^{-2\pi i \mathbf{r} \cdot \hat{\mathbf{s}}}\, d\Omega
\end{equation}  
By making the latter assumptions we have left treatment of the ionosphere and terrestrial 
emission such as local interference or thermal radiation from the ground for future work. }

\textbf{Next we turn to the measurement performed by an antenna with far-field radiation pattern 
$\mathcal{W}_a(\hat{\mathbf{s}})$, where the subscript $a$ denotes a specific antenna. We 
adopt the common convention of normalizing the radiation pattern to have response of 1 at the 
center of the main lobe \citep[e.g.][]{nap99}. Consequently, the translation between the terminal 
voltages to electric field measurements has been absorbed in our definition of the radiation 
pattern. The ideal antenna electric field measurement can be expressed as 
\begin{equation}
E^I_a(f,t) = \int \mathcal{W}_a(\hat{\mathbf{s}})\mathcal{E}(\hat{\mathbf{s}},f,t) e^{-2\pi i \mathbf{r} \cdot \hat{\mathbf{s}}}\, d\Omega. 
\end{equation}
Equivalently we can express the effect of the antenna radiation pattern in the plane of the 
aperture.
\begin{equation}
E^I_a(f,t) = \int_{\mathrm{aperture}}W_a(\mathbf{r}-\mathbf{r}_a)E_a(\mathbf{r},f,t)\,d^2\mathbf{r}
\end{equation}
Here we used the Fourier relationship between the antenna's far-field radiation pattern and the 
aperture illumination pattern (sometimes referred to as the aperture distribution), $W_a(\mathbf{r}-\mathbf{r}_a)=\int  \mathcal{W}_a(\hat{\mathbf{s}}) \, d
\Omega$. 
While it is possible to define each antenna aperture illumination pattern in a way to absorb the 
relative position of the antenna, we keep the $\mathbf{r}_a$ coordinate to retain the separable
concepts of the primary beam (the product of two antenna radiation patterns) and the array
point spread function (square of the Fourier transform of the array layout).}

\textbf{The first step of the MOFF algorithm is to digitize the antenna signals and perform an FFT to 
channelize the time-ordered data. This is effectively estimating the quasi-monochromatic 
components of the electric field we already discussed. The sampling rate of the digitization 
determines the bandwidth of the observation, and the sampling period determines the minimum 
frequency channel resolution. The frequency Fourier transform operation is often referred to as 
the F-engine of a correlator. The MOFF shares the same F-engine design as an FX cross
correlator which performs the frequency transform (F-engine) before cross multiplying 
(X-engine). }

Thus far we have \textbf{considered the perfectly measured electric field distribution according to 
the antenna far-field radiation patterns}. Here we introduce a multiplicative complex gain as 
well as an additive noise term which corrupt the measured signals.
\begin{equation}\label{eq:apply_gain}
E_a(f,t) = g_a(f,t) E_a^I(f,t) + n_a(f,t)
\end{equation}
\textbf{We have carried the frequency dependence to this point to make clear the complex gains are
frequency dependent. However,} each subsequent step treats each frequency 
channel independently, so we will drop the $f$ to simplify notation. 
\textbf{Furthermore, while antenna gains may vary with time, we will assume they are constant over
the period of our calibration, and thus drop their time dependence from the notation.}

The MOFF algorithm next calls for a calibration \textbf{of the electric field measurements}. The goal of our new calibration method will be 
to form an estimate of the gains, $g'_a$, \textbf{where the prime represents an estimate}. For now we will assume we have formed an estimate 
to proceed with the MOFF pipeline. We correct the incoming electric field data stream using our 
current estimate of the gains.
\begin{equation}
E'_a(t) = E_a(t)/ g'_a
\end{equation}

The next step is to grid the \textbf{calibrated electric field measurements to a finite set of regularly spaced 
positions on the 2D plane of the ground, $\mathbf{r}_j$.
The extent of the grid is chosen to encompass the footprint of the antenna array.
MOFF uses the antenna aperture illumination patterns as the} gridding 
kernel according to the software holography/A-transpose technique \citep{mor09,bha08}. 
\textbf{Mathematically we can express the gridding operation as}
\begin{equation}\label{eq:gridding}
E'_j=\sum_a W_a(\mathbf{r}_j-\mathbf{r}_a) E'_a(t)
\end{equation}
\textbf{Formally each pixel is calculated by summing over all antennas and their aperture illumination patterns.
However the aperture illumination patterns are typically compact and non-overlapping, so most pixels
will only contain contribution from at most one antenna, allowing equation~\ref{eq:gridding}
to be calculated through a sparse matrix multiplication.}
This gridding 
achieves an optimal map \citep{teg97b} and places the data on a regular grid. Unlike 
traditional correlators which grid visibilities, the MOFF grids the electric field measurements directly. This 
operation must be performed for every time sample, before any averaging of the data.

After gridding, the imaging portion of the MOFF performs a 2D spatial FFT to form 
instantaneous electric field images of the sky \textbf{at fixed locations $\s_k$.
The Fourier dual coordinates to $\mathbf{r}$ are $(l,m)$, where $l=\sin\theta\cos\phi$,
$m=\nobreak\sin\theta\sin\phi$, and $(\theta,\phi)$ are the zenith 
and azimuthal angles. The full unit vector towards the sky is 
$\s=(l,m,\sqrt{1-l^2-m^2})$.
The sample spacing and resolution of $\s_k$ are given 
by the choices of $\mathbf{r}_j$ grid parameters, according to the usual FFT relations. 
Fine $\mathbf{r}_j$ grid spacing can result in non-physical angular pixels ($l^2+m^2>1$),
which should be discarded.}

The pipeline up to this point can be summarized 
through the following equation.
\begin{align}
\mathcal{E}'(\s_k,t) & = \underbrace{\frac{1}{\Nant} \sum_j e^{2\pi i \mathbf{r}_j \cdot \s_k}}_{\mathrm{2D\;FFT}} 
\underbrace{\sum_a W_a(\mathbf{r}_j - \ra)}_{\mathrm{Gridding}} \\
& \qquad \times \underbrace{ \frac{1}{g'_a} \left(g_a E_a^I(t)+n_a(t)\right)}_{\mathrm{Calibration}} \nonumber
\end{align}
The sum over antennas uses the \textbf{antenna aperture illumination pattern} to grid the calibrated electric field \textbf{measurements} onto 
regular gridpoints, $\mathbf{r}_j$. The \textbf{sum over $j$} denotes the 2D FFT to sky coordinates, resulting in an 
estimate for the instantaneous electric field image. We can simplify this expression by 
exchanging the sums \textbf{and use the DFT to approximate the Fourier relationship between
the antenna aperture illumination pattern and the far-field radiation pattern.}
\begin{align}\label{eq:epix}
\mathcal{E}'(\s_k,t) & = \frac{1}{\Nant} \sum_a \frac{1}{g'_a}\left(g_a E^I_a(t)+n_a(t)\right) e^{2\pi i \s_k \cdot \ra} \nonumber \\
  & \qquad\times \sum_j W_a(\mathbf{r}_j-\ra)e^{2\pi i \s_k \cdot (\mathbf{r}_j-\ra)} \nonumber\\
& \approx \frac{1}{\Nant} \sum_a \frac{1}{g'_a}\left(g_aE^I_a(t)+n_a(t)\right) e^{2\pi i \s_k \cdot \ra}\mathcal{W}_a(\s_k)
\end{align}
In this form we see that the effect of gridding with the antenna \textbf{aperture illumination pattern} is to attenuate 
the image by a factor of the \textbf{far-field radiation response, $\mathcal{W}_a(\s)$.}

Finally, the images are squared and averaged in time to form \textbf{dirty images, 
$I'(\hat{\mathbf{s}}_k)$, then} written 
to disk. While operations have been reordered, the resulting image is equivalent to one 
produced by gridding and imaging visibilities. By shifting from cross-correlation of all pairs of 
antennas to a spatial FFT, the computational cost is significantly lowered for certain classes of 
compact arrays. The output bandwidth is also greatly reduced as visibilities are exchanged for 
gridded images. A full exploration of these scalings was presented in \citealt{thy15c}.

@@

As a starting point, we consider the feedback calibration outlined in \citealt{mor11}. There it 
was suggested to form a correlation of the uncalibrated antenna measurements with the \textbf{complex conjugate of} an image 
pixel from the output of the correlator. This is a statistically stationary quantity, which can be 
related to the sum of visibilities involving the antenna used in the correlation -- exactly the sum 
needed to calibrate a simple sky of a single point source. However, we aim for a more 
generalized solution for arbitrarily complex sky models. We therefore study the full expression 
for the antenna-pixel correlation,
\begin{equation}\label{eq:Kna_def}
\Kna \equiv \left<E_a^{(n)}(t) \mathcal{E}'^{*(n)}(\spix,t)\right>_t,
\end{equation}
where the superscript $n$ again represents the quantity formed in the $n^\mathrm{th}$ 
calibration loop, 
\textbf{the superscript $^*$ indicates complex conjugation,}
and \spix\, is the pixel center nearest a bright calibrator of interest. The 
following will hold for any chosen pixel, \spix, though it is advantageous to choose a pixel which 
contains a bright source to achieve a high signal to noise.

Plugging equation \ref{eq:epix} into equation \ref{eq:Kna_def}, 
\textbf{and substituting the $g^{(n)}_b$ for $g'_b$ to indicate the specific loop number,}
we find,
\begin{align}
\Kna & = \left<\left(g_a E_a^I(t)+n_a(t)\right) \frac{1}{\Nant} \right. \nonumber \\
& \qquad \left. \times \sum_b \frac{1}{g^{*(n)}_b}\left(g^*_bE_b^{I*}(t)+n^*_b(t)\right) e^{-2\pi i \spix \cdot \rb} \mathcal{W}^*_b(\spix)\right>_t \nonumber \\
& = \frac{1}{\Nant} \sum_b \frac{1}{g^{*(n)}_b} \mathcal{W}^*_b(\spix) e^{-2\pi i \spix \cdot \rb} \nonumber \\
& \qquad \times \left<\left(g_aE_a^I(t)+n_a(t)\right)\left(g^*_bE_b^{I*}+n^*_b(t)\right) \right>_t \nonumber \\
& = \frac{1}{\Nant} \sum_b \frac{1}{g^{*(n)}_b} \mathcal{W}^*_b(\spix) e^{-2\pi i \spix \cdot \rb} \nonumber \\
& \qquad \times \left(g_a g^*_b \left<E^I_a(t)E^{*I}_b(t)\right>_t + g_a\left<E_a^I(t)n^*_b(t)\right>_t \right. \nonumber \\
& \qquad \left. + g^*_b\left<n_a(t)E^{I*}_b(t)\right>_t + \left<n_a(t) n^*_b(t)\right>_t\right)  
%& = \frac{1}{\Nant} \sum_b \frac{1}{g^{*(n)}_b} \mathcal{W}^*_b(\spix) e^{-2\pi i \spix \cdot \rb} \left(g_ag^*_bV^I_{ab}+n_{ab}\right)
\end{align}
where in the second step we group time-dependent terms, 
and in the third \textbf{step we expanded the product within the time average.
Because the noise does not correlate with the ideal antenna measurements we can
drop the middle terms. We will also assume that the cross-correlation noise ($a\ne b$) 
is zero mean. Then by defining an ideal visibility as
$V_{ab}^I \equiv \left<E^I_a(t)E^{*I}_b(t)\right>_t$, we can simplify to}
\begin{align}\label{eq:Kna}
\Kna & = \frac{1}{\Nant} \sum_b \frac{1}{g^{*(n)}_b} \mathcal{W}^*_b(\spix) e^{-2\pi i \spix \cdot \rb}\nonumber \\
& \qquad \times \left( g_ag_b^* V_{ab}^I + \delta_{ab}\left<|n_a(t)|^2\right>_t\right)
\end{align}
\textbf{where $\delta_{ab}$ is the Kronecker-delta function. To be clear, visibilities are never directly 
formed from the data. Rather it is convenient to express $\Kna$ in terms of the ideal 
visibilities, which we can model when solving for the gains below.}
